---
title: "Predicting Profitability, Returns, and Customer Value in Shopify Sales"
subtitle: "Final Report"
author: "Group 2: <br> Kevin Bachez, Alexa Bueno, Leeann Lewis-Waddell, <br> Ron Stephen Sabilla, Brian Vuong"
format:
  revealjs:
    theme: [default, shopifytheme5.scss]
    slide-level: 1
    center-title-slide: true
    title-slide-attributes:
      style: "font-size: 1.6em; height: 100%; min-height: 100%; box-sizing: border-box;"
    slide-number: true
    controls: true
    width: 2500
    height: 1000
    margin: 0.08
    min-scale: 0.2
    max-scale: 1
footer: "IBM6540 | August 12, 2026"
editor: 
  markdown: 
    wrap: 72
---

# Table of Contents

:::::: columns
::: {.column width="30%"}
[**Introduction**]{.underline}

-   Problem definition

-   Research objectives

-   Importance of the research

-   Population / Sample

-   Sample Characteristics
:::

::: {.column width="30%"}
[**Research**]{.underline}

-   Analytics Objective 1

-   Analytics Objective 2

-   Analytics Objective 3

-   Analytics Objective 4

-   Analytics Objective 5
:::

::: {.column width="30%"}
[**Conclusion**]{.underline}

-   Key Findings

-   Recommendations for Businesses

-   Limitations and Directions for Future Research
:::
::::::

# Introduction

## Problem Definition

------------------------------------------------------------------------

## The Core Challenge

-   **The Issue:** Top-line gross transaction value is growing, but net
    profit margins are being aggressively eroded
-   **The Friction:** High acquisition costs from `Paid Ads` &
    `Social Media` compounded by excessive discounting and high return
    rates
-   **The Gap:** No automated system to identify which product category,
    traffic source, and discount combinations trigger unprofitable
    transactions

------------------------------------------------------------------------

## Business & Financial Impact

-   **Margin Compression:** Heavy discounts (`discount_percent`) drive
    volume (`quantity`) but cannibalize net `profit` per order

-   **Logistical Drain:** Returns (`is_returned`) create operational
    bottlenecks and sunk costs through absorbed `shipping_cost` overhead

-   **The Goal:** Build an ML framework to optimize marketing spend,
    predict return probability, and maximize net profit margins

------------------------------------------------------------------------

# Research Objectives

- **RO1 – Marketing Channels & Average Order Value:** Determine which marketing traffic source generates the highest average order value and whether this relationship differs across product categories.

- **RO2 – Customer Satisfaction Analysis:** Identify which customer and transaction characteristics contribute to higher customer satisfaction by examining how product categories, traffic sources, discount levels, and purchasing behavior influence customer rating.

- **RO3 – Returns & Profitability Analysis:** Evaluate how product categories, shipping costs, customer ratings, and discount structures impact the probability of an item being returned and quantify the resulting drain on net profit.

- **RO4 – Profit Impact of Customer Retention:** Determine whether repeat customers are a more margin-efficient growth channel than new customer acquisition.

- **RO5 – Sales Trends & Seasonality:** Determine how sales performance changes over time and whether seasonal patterns differ across product categories.

# Importance of the Research

::: incremental
-   Top-line transaction value is growing across the business, but **net
    profit margins are being eroded** — and no single team or department
    owns the full explanation why
-   Acquisition costs, discounting behavior, customer satisfaction,
    return rates, and retention all interact, but are typically analyzed
    in isolation — this project connects them
-   Understanding *which* marketing channels perform best (AO1), what
    drives customer satisfaction (AO2), what triggers returns (AO3),
    whether repeat customers are more profitable (AO4), and how sales
    shift seasonally (AO5) gives a full-funnel view of where margin is
    actually being lost
-   Predictive models across these objectives let the business move from
    reactive reporting to **proactively flagging unprofitable
    channel–category–discount combinations before they happen**
-   This addresses a real operational gap: the business currently lacks
    any automated, predictive system tying acquisition, satisfaction,
    returns, and retention together
:::

# Population / Sample

## Shopify e-commerce orders from January 2023 to June 2025

-   **Unit of analysis:** Individual order (n = 60,000)

-   31,154 unique customers \| 6,998 unique products

-   **Scope limit:** Findings generalize only to these countries,
    categories, and date range

|              |             |        |             |          |        |         |            |
|---------|---------|---------|---------|---------|---------|---------|---------|
| **Country**  | USA         | UAE    | CAN         | UK       | IND    | GER     | AUS        |
| **Category** | Accessories | Sports | Electronics | Footwear | Beauty | Fashion | Home Decor |

[Kaggle, Shopify Sales Dataset for ML &
EDA](https://www.kaggle.com/datasets/aliiihussain/shopify-sales-dataset-for-ml-and-eda/data)

::: callout-important
Kaggle notes the dataset is "realistic," presumably meaning transactions
are not verified, and thus, **synthetic**.
:::

# Sample Characteristics

-   The dataset contains **60,000 Shopify e-commerce orders** from
    January 2023 to June 2025

-   The sample includes **31,154 unique customers** and **6,998 unique
    products**

-   Customers are located across **7 countries**: USA, UAE, Canada, UK,
    India, Germany, and Australia

-   Products are grouped into **7 categories**: Accessories, Sports,
    Electronics, Footwear, Beauty, Fashion, and Home Decor

-   Key variables capture **customer characteristics, product
    information, pricing, discounts, traffic sources, ratings, returns,
    and purchasing behavior**

-   This allows the analysis to examine the customer journey across
    **acquisition, satisfaction, returns, profitability, retention, and
    seasonality** **:::**

# Analytics Objective 1

Predict **average order value (AOV)** from traffic source & product
category. Identify which **channel × category combination** drives the
highest order value.

------------------------------------------------------------------------

## Machine Learning Methods

-   **Regression:** Linear, Ridge, Lasso, Elastic Net
-   **Classification:** Logistic Regression, KNN, Decision Tree *(AOV →
    Low/Med/High)*
-   **Ensembles:** Random Forest, XGBoost, Bagged Trees

------------------------------------------------------------------------
```{r ao1-setup}
#| include: false

library(tidyverse)
library(tidymodels)
library(janitor)
set.seed(2024)

df_raw <- read_csv("shopify_sales_dataset_ml_eda 2.csv", show_col_types = FALSE)
df     <- df_raw |> clean_names()

df_ro1 <- df |>
  select(revenue, traffic_source, product_category) |>
  filter(!is.na(revenue), !is.na(traffic_source),
         !is.na(product_category), revenue > 0) |>
  mutate(
    traffic_source   = factor(traffic_source),
    product_category = factor(product_category)
  )
```

## Data Wrangling & Visualization

- Filter missing values; keep only valid orders (revenue > 0)
- One-hot encode channel & category; bin AOV into tiers
- Boxplots by channel, heatmap of mean AOV across channel × category

```{r ao1-heatmap}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

df_ro1 |>
  group_by(traffic_source, product_category) |>
  summarise(mean_aov = round(mean(revenue)), .groups = "drop") |>
  ggplot(aes(x = traffic_source, y = product_category, fill = mean_aov)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = scales::dollar(mean_aov)), size = 3.5, color = "white") +
  scale_fill_gradient(low = "#1a5276", high = "#b7950b", labels = scales::dollar) +
  labs(title = "Mean AOV: Traffic Source × Product Category",
       x = "Traffic Source", y = "Product Category", fill = "Mean AOV ($)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

------------------------------------------------------------------------

## Model Fitting

-   80/20 stratified split + 10-fold CV
-   Recipe: normalize, one-hot encode, bin AOV for classification
-   Tune: penalty, mtry, trees, learn rate, tree depth

------------------------------------------------------------------------

## Prediction & Model Evaluation

- **Regression:** RMSE · R² via 5-fold cross-validation
- Linear Regression vs. Random Forest evaluated on the same folds
- Best model selected by lowest RMSE; evaluated on held-out test set

```{r ao1-eval}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| cache: true

set.seed(2024)
ao1_split <- initial_split(df_ro1, prop = 0.80, strata = revenue)
ao1_train <- training(ao1_split)
ao1_test  <- testing(ao1_split)
ao1_folds <- vfold_cv(ao1_train, v = 5, strata = revenue)

base_rec <- recipe(revenue ~ traffic_source + product_category, data = ao1_train) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

lm_res <- workflow() |>
  add_recipe(base_rec) |>
  add_model(linear_reg() |> set_engine("lm")) |>
  fit_resamples(resamples = ao1_folds, metrics = metric_set(rmse, rsq)) |>
  collect_metrics() |> mutate(model = "Linear Regression")

rf_res <- workflow() |>
  add_recipe(base_rec) |>
  add_model(rand_forest(trees = 100) |> set_engine("ranger") |> set_mode("regression")) |>
  fit_resamples(resamples = ao1_folds, metrics = metric_set(rmse, rsq)) |>
  collect_metrics() |> mutate(model = "Random Forest")

bind_rows(lm_res, rf_res) |>
  filter(.metric == "rmse") |>
  ggplot(aes(x = reorder(model, mean), y = mean, fill = model)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                width = 0.3, color = "#b7950b", linewidth = 0.8) +
  coord_flip() +
  labs(title = "5-Fold CV RMSE — Model Comparison",
       subtitle = "Lower RMSE = better | Error bars = ± 1 SE",
       x = NULL, y = "Mean CV RMSE ($)") +
  scale_fill_manual(values = c("Linear Regression" = "#1a5276",
                               "Random Forest"     = "#b7950b")) +
  theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

## Interpretation

- **product_category_Electronics** and **traffic_source_Social.Media** are the top AOV-driving features
- Social Media × Accessories drives the highest mean AOV (~$1,039) per the heatmap
- Channel and category together explain consistent AOV patterns across all 60,000 orders

```{r ao1-vip}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| cache: true

library(vip)

rf_spec <- rand_forest(trees = 100) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

rf_rec <- recipe(revenue ~ traffic_source + product_category, data = df_ro1) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

rf_fit <- workflow() |>
  add_recipe(rf_rec) |>
  add_model(rf_spec) |>
  fit(data = df_ro1)

rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 10,
      aesthetics = list(fill = "#1a5276", alpha = 0.85)) +
  labs(title = "Feature Importance — Random Forest (AOV)") +
  theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

## Summary Findings

- **Social Media × Accessories** = highest mean AOV (~$1,039) across all channel × category combinations
- **Electronics** and **Social Media** are the top AOV-driving features per Random Forest variable importance
- Linear Regression and Random Forest achieve similar RMSE (~$800) → channel × category relationship is largely linear
- Prioritize marketing spend on **Social Media → Electronics & Accessories** to maximize revenue per order

# Analytics Objective 2

## Customer Satisfaction

**Objective:** Identify which customer and transaction characteristics contribute to higher customer satisfaction by examining how product categories, traffic sources, discount levels, and purchasing behavior influence customer `rating`.

### Variables

- **Dependent Variable:** `rating`
- **Independent Variables:** `product_category`, `traffic_source`, `discount_percent`, and `quantity`


### Methods

Two regression approaches were used to evaluate customer satisfaction:

- Multiple Linear Regression
- Random Forest Regression
- 80/20 train/test split
- RMSE, MAE, and R² for model evaluation

```{r ao2-setup}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(janitor)
library(ranger)
library(broom)
library(knitr)

set.seed(2026)

ao2_raw <- read_csv(
  "shopify_sales_dataset_ml_eda 2.csv",
  show_col_types = FALSE
) |>
  clean_names()

ao2_data <- ao2_raw |>
  select(
    rating,
    product_category,
    traffic_source,
    discount_percent,
    quantity
  ) |>
  filter(
    !is.na(rating),
    !is.na(product_category),
    !is.na(traffic_source),
    !is.na(discount_percent),
    !is.na(quantity)
  ) |>
  mutate(
    product_category = factor(product_category),
    traffic_source = factor(traffic_source)
  )
```

---

## Customer Ratings by Traffic Source

Customer ratings were compared across traffic sources to determine whether customers from different marketing channels showed differences in satisfaction.

```{r ao2-traffic-rating}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| message: false
#| warning: false

ggplot(
  ao2_data,
  aes(
    x = traffic_source,
    y = rating
  )
) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "Customer Ratings by Traffic Source",
    x = "Traffic Source",
    y = "Customer Rating"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )
```

The distributions were relatively similar across traffic sources, suggesting that marketing channel was not strongly associated with customer satisfaction.

---

## Model Development

An 80/20 train/test split was used to compare Multiple Linear Regression and Random Forest Regression.

```{r ao2-model-fitting}
#| include: false
#| message: false
#| warning: false

set.seed(2026)

ao2_split <- initial_split(
  ao2_data,
  prop = 0.80,
  strata = rating
)

ao2_train <- training(ao2_split)
ao2_test <- testing(ao2_split)

ao2_recipe <- recipe(
  rating ~
    product_category +
    traffic_source +
    discount_percent +
    quantity,
  data = ao2_train
) |>
  step_dummy(
    all_nominal_predictors(),
    one_hot = TRUE
  ) |>
  step_zv(all_predictors())

ao2_lm_spec <- linear_reg() |>
  set_engine("lm")

ao2_rf_spec <- rand_forest(
  trees = 500,
  mtry = 4,
  min_n = 5
) |>
  set_engine(
    "ranger",
    importance = "impurity"
  ) |>
  set_mode("regression")

ao2_lm_wf <- workflow() |>
  add_recipe(ao2_recipe) |>
  add_model(ao2_lm_spec)

ao2_rf_wf <- workflow() |>
  add_recipe(ao2_recipe) |>
  add_model(ao2_rf_spec)

ao2_lm_fit <- fit(
  ao2_lm_wf,
  data = ao2_train
)

ao2_rf_fit <- fit(
  ao2_rf_wf,
  data = ao2_train
)

ao2_lm_predictions <- augment(
  ao2_lm_fit,
  new_data = ao2_test
)

ao2_rf_predictions <- augment(
  ao2_rf_fit,
  new_data = ao2_test
)

ao2_metrics_function <- metric_set(
  rmse,
  mae,
  rsq
)

ao2_metrics <- bind_rows(

  ao2_lm_predictions |>
    ao2_metrics_function(
      truth = rating,
      estimate = .pred
    ) |>
    mutate(model = "Linear Regression"),

  ao2_rf_predictions |>
    ao2_metrics_function(
      truth = rating,
      estimate = .pred
    ) |>
    mutate(model = "Random Forest")

) |>
  select(
    model,
    .metric,
    .estimate
  )
```

---

## Model Performance

The two models produced nearly identical prediction accuracy.

```{r ao2-model-table}
#| echo: false
#| message: false
#| warning: false

ao2_metrics |>
  mutate(
    .estimate = round(.estimate, 3)
  ) |>
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) |>
  select(
    model,
    rmse,
    mae,
    rsq
  ) |>
  knitr::kable(
    col.names = c(
      "Model",
      "RMSE",
      "MAE",
      "R²"
    ),
    caption = "Customer Rating Model Performance"
  )
```

```{r ao2-model-comparison}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| message: false
#| warning: false

ao2_metrics |>
  filter(.metric == "rmse") |>
  ggplot(
    aes(
      x = model,
      y = .estimate
    )
  ) +
  geom_col(
    fill = "steelblue",
    width = 0.6
  ) +
  geom_text(
    aes(label = round(.estimate, 3)),
    vjust = -0.5,
    size = 4
  ) +
  labs(
    title = "Linear Regression vs. Random Forest",
    subtitle = "Lower RMSE indicates better predictive performance",
    x = NULL,
    y = "RMSE"
  ) +
  theme_minimal()
```

The Linear Regression model achieved an RMSE of approximately **1.15**, while the Random Forest model produced nearly identical performance. Both models had an R² of approximately **0**, meaning that the selected predictors explained very little variation in customer ratings.

---

## Summary Findings

- Product category, traffic source, discount percentage, and quantity did **not show a strong relationship with customer ratings**.
- Linear Regression and Random Forest produced nearly identical results, with an RMSE of approximately **1.15**.
- Both models had an **R² of approximately 0**, indicating that the selected variables explained very little variation in customer satisfaction.
- None of the tested predictors were statistically significant.
- Overall, customer satisfaction may be influenced more by factors not included in the model, such as **product quality, delivery experience, or customer service**.

# Analytics Objective 3

Predict **return risk (`is_returned`)** from category, discount, shipping cost & rating, and assess the downstream profit impact of high-risk orders.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(discrim)
library(DALEX)
library(DALEXtra)
library(xgboost)

raw <- read_csv("shopify_sales_dataset_ml_eda 2.csv")


tidymodels_prefer()
```

---

## Machine Learning Methods

- **Regression:** Logistic Regression *(binary outcome)*
- **Classification:** LDA, QDA, KNN, Decision Tree
- **Ensembles:** Bagged Trees, Random Forest, XGBoost

---

## Data Wrangling & Visualization

```{r}
#| label: load-wrangle-data
#| echo: false
#| message: false
#| warning: false

df <- read_csv("shopify_sales_dataset_ml_eda 2.csv") %>%
  mutate(
    is_returned      = factor(is_returned, levels = c(0, 1), labels = c("No", "Yes")),
    product_category = as.factor(product_category)
  )
```

```{r}
#| label: return-rate-by-category
#| echo: false

df %>%
  group_by(product_category) %>%
  summarise(return_rate = mean(is_returned == "Yes")) %>%
  ggplot(aes(x = reorder(product_category, -return_rate), y = return_rate)) +
  geom_col(fill = "steelblue") +
  labs(x = "Product Category", y = "Return Rate", title = "Return Rate by Category") +
  theme_minimal() +
  coord_flip()
```

- Convert `is_returned` & `product_category` to factors; confirm zero missing values
- Stratify splits/CV on `is_returned` — ~15% return rate (class imbalance)

---

## Model Fitting

```{r}
#| label: fit-model
#| echo: false
#| message: false
#| warning: false

set.seed(123)
data_split <- initial_split(df, prop = 0.75, strata = is_returned)
train_data <- training(data_split)
test_data  <- testing(data_split)

returns_recipe <- recipe(
    is_returned ~ product_category + discount_percent + shipping_cost + rating,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

boost_spec <- boost_tree(trees = 50) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

boost_wf  <- workflow() %>% add_recipe(returns_recipe) %>% add_model(boost_spec)
boost_fit <- fit(boost_wf, data = train_data)
```

- 75/25 stratified split + 3–10 fold CV
- Recipe: dummy-encode categoricals, zero/near-zero variance filter *(no normalization for trees)*
- Tune: cost complexity, mtry, trees, learn rate, neighbors (k)

---

## Prediction and Model Evaluation

- **Primary metric:** ROC AUC, 10-fold CV mean *(class imbalance makes accuracy misleading)*
- Confusion Matrix · Sensitivity/Specificity per model
- Best model by CV `roc_auc` → **XGBoost** *(margin over runner-up discussed in Summary Findings)*

---

## Interpretation

```{r}
#| label: shap-generation
#| echo: false
#| message: false
#| warning: false

xgb_engine <- extract_fit_engine(boost_fit)

set.seed(123)
shap_sample <- train_data[sample(nrow(train_data), 50), ]

xgb_explainer <- explain_tidymodels(
  boost_fit,
  data = shap_sample %>% select(product_category, discount_percent, shipping_cost, rating),
  y = as.numeric(shap_sample$is_returned) - 1,
  label = "XGBoost",
  verbose = FALSE
)

shap_results <- purrr::map_dfr(1:nrow(shap_sample), function(i) {
  sp <- predict_parts(xgb_explainer, new_observation = shap_sample[i, ], type = "shap", B = 25)
  as.data.frame(sp) %>% mutate(obs_id = i)
})

shap_summary <- shap_results %>%
  group_by(variable_name) %>%
  summarise(mean_abs_shap = mean(abs(contribution)), .groups = "drop") %>%
  arrange(desc(mean_abs_shap))
```

```{r}
#| label: shap-summary
#| echo: false

ggplot(shap_summary, aes(x = reorder(variable_name, mean_abs_shap), y = mean_abs_shap)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Mean |SHAP Value| by Feature", x = "Feature", y = "Mean |SHAP value|") +
  theme_minimal()
```

```{r}
#| label: variable-importance
#| echo: false

importance_matrix <- xgboost::xgb.importance(model = xgb_engine)

ggplot(importance_matrix, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(title = "XGBoost Variable Importance (Gain)", x = "Feature", y = "Gain") +
  theme_minimal()
```

- SHAP (`predict_parts()`) → `shipping_cost` ranks #1, ahead of `discount_percent` *(ranking taken with the predictive-strength caveat in Summary Findings)*
- Variable importance (manual `ggplot2`, `xgb.importance()`) confirms SHAP ranking
- Framing: discount % & shipping cost = pre-purchase levers; rating = post-purchase signal

---

## Summary Findings

- **XGBoost** = top model by 10-fold CV, ROC AUC = **0.509** (SE ≈ 0.003) — narrowly ahead of Lasso Logistic Regression (0.505), Decision Tree (0.503), and Random Forest (0.501), though the margin is within one standard error of the runner-up, so the lead is not statistically decisive
- **Shipping cost**, not discount or category, is the leading return-risk driver → diverges from original hypothesis
- SHAP & PDPs confirm shipping cost and discount % as key, actionable drivers
- Deployed as a self-contained client-side calculator for pre-purchase return-risk scoring
- Profit impact: On average, a **returned order carries ~$87 in sunk shipping and discount cost** (avg. shipping cost ≈ \$13.44 + avg. discount amount ≈ \$73.74) that is spent regardless of the return, with **no offsetting revenue retained**. Applied across the 8,887 returned orders (14.8% of the 60,000-order dataset), this represents roughly **\$775K in aggregate sunk cost** tied to returns. Notably, raw profit-per-order is nearly identical between returned (\$974.81) and kept (\$981.58) orders — the ~\$6.77 gap is small, reinforcing that return risk in this dataset is not driven by a large per-order profit penalty on its own, but by the compounding volume of returns at scale. This closes the loop on the opening objective: even a modest per-order sunk cost becomes a meaningful margin drain in aggregate, which is exactly the kind of pattern a pre-purchase risk-scoring calculator (Section 2.2) is designed to intercept before shipping and discount costs are locked in.

::: {.callout-note title="On model selection"}
Per the decision rule stated earlier in this document (@sec-cv-comparison), a lead smaller than one SE would favor **Lasso LR** on parsimony grounds. XGBoost is retained here as the deployed model primarily for its PDP/SHAP interpretability story and consistency with the group's established modeling narrative, not because its CV performance is meaningfully superior.
:::

::: {.callout-warning title="On predictive strength"}
An ROC AUC of ~0.51 is only marginally above chance (0.50), meaning none of
the four candidate models — including XGBoost — meaningfully separate
returned from non-returned orders using this feature set alone. The SHAP
ranking (`shipping_cost` > `discount_percent` > `rating` > `product_category`)
should be read as a relative signal within a weak model, not as strong
evidence that shipping cost drives returns in practice. This constrains how
confidently the client recommendations above can be stated and points to the
feature-set limitation noted in the group's Limitations & Future Research
section.
:::

# Analytics Objective 4

Predict whether a new customer will become **high-value** (top-quartile
lifetime profit) from their first order's `product_price` and
`discount_percent`, so the business can flag high-potential customers
before their full purchase history accumulates.

# Machine Learning Methods

-   Classification: LDA (primary), QDA & Naïve Bayes tested for
    comparison
-   Ensembles: Random Forest
-   Regression: not used — outcome is binary (`high_value`), not
    continuous

# Data Wrangling & Visualization

-   Aggregate to customer level; derive `cumulative_profit` = sum of
    `profit` per `customer_id`
-   Label `high_value` = 1 if `cumulative_profit` ≥ 75th percentile
    (\$2,698.35) → 7,789/31,154 customers (\~25%)
-   Isolate each customer's first order (`arrange` + `slice(1)`) for
    `product_price`, `discount_percent` — no post-purchase leakage
-   `product_category` and acquisition `traffic_source` tested and
    dropped (ANOVA p = 0.41, p = 0.15 — no significant effect on profit)
-   Scatterplot: price vs. cumulative profit, colored by `high_value` ·
    Boxplot: discount % by `high_value` status

# Data Wrangling & Visualization

```{r}
#| label: fig-ao4-wrangling
#| fig-cap: "First-order price and discount by customer value"

library(tidymodels)
library(tidyverse)

raw <- read_csv("shopify_sales_dataset_ml_eda 2.csv")

customer_profit <- raw |>
  group_by(customer_id) |>
  summarise(cumulative_profit = sum(profit), .groups = "drop") |>
  mutate(high_value = factor(if_else(
    cumulative_profit >= quantile(cumulative_profit, 0.75), 1, 0
  ), levels = c(1, 0)))

first_orders <- raw |>
  arrange(customer_id, order_date, order_id) |>
  group_by(customer_id) |>
  slice(1) |>
  ungroup() |>
  select(customer_id, product_price, discount_percent)

model_data <- customer_profit |>
  left_join(first_orders, by = "customer_id") |>
  select(customer_id, high_value, product_price, discount_percent, cumulative_profit)

p1 <- ggplot(model_data, aes(x = product_price, y = cumulative_profit, color = high_value)) +
  geom_point(alpha = 0.3) +
  labs(title = "First-Order Price vs. Cumulative Profit", x = "Price ($)", y = "Cumulative Profit ($)", color = "High Value") +
  theme_minimal(base_size = 11)

p2 <- ggplot(model_data, aes(x = high_value, y = discount_percent, fill = high_value)) +
  geom_boxplot() +
  labs(title = "First-Order Discount by Customer Value", x = "High Value", y = "Discount (%)") +
  theme_minimal(base_size = 11)

library(patchwork)
p1 + p2
```

# Model Fitting

-   80/20 stratified split on `high_value` (class balance preserved by
    construction, \~25/75)
-   Recipe: `step_normalize()` on both predictors — no dummy step needed
    (no categoricals retained)
-   No tuning parameters for LDA; `mtry`/`trees` left at tidymodels
    defaults for Random Forest (baseline comparison only, not the
    deployed model)

# Model Fitting

```         
library(discrim)
library(ranger)

set.seed(2025)
  split <- initial_split(model_data, prop = 0.80, strata = high_value)
  train_data <- training(split)
  test_data  <- testing(split)

hv_recipe <- recipe(high_value ~ product_price + discount_percent + customer_id, data = train_data) |>
  update_role(customer_id, new_role = "ID") |>
  step_normalize(all_numeric_predictors())

lda_wf <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(discrim_linear() |> 
  set_engine("MASS") |> 
  set_mode("classification"))
  
rf_wf  <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(rand_forest(trees = 300) |> 
  set_engine("ranger", importance = "impurity") |> 
  set_mode("classification"))

lda_fit <- fit(lda_wf, data = train_data)
rf_fit  <- fit(rf_wf,  data = train_data)
```

# Prediction & Model Evaluation

-   Primary metric: ROC AUC on held-out test set
-   LDA AUC = **0.679** vs. Random Forest AUC = **0.590** → LDA selected
    (added tree complexity not justified; relationship is linear)
-   At default 0.5 cutoff, high-value recall is low (\~5%) due to class
    imbalance — deployed version scores by **probability**, not hard
    class, so the business sets its own targeting threshold (e.g., top
    20%)

# Prediction & Model Evaluation

```{r}

library(discrim)
library(ranger)

set.seed(2025)
  split <- initial_split(model_data, prop = 0.80, strata = high_value)
  train_data <- training(split)
  test_data  <- testing(split)

hv_recipe <- recipe(high_value ~ product_price + discount_percent + customer_id, data = train_data) |>
  update_role(customer_id, new_role = "ID") |>
  step_normalize(all_numeric_predictors())

lda_wf <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(discrim_linear() |> 
  set_engine("MASS") |> 
  set_mode("classification"))
  
rf_wf  <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(rand_forest(trees = 300) |> 
  set_engine("ranger", importance = "impurity") |> 
  set_mode("classification"))

lda_fit <- fit(lda_wf, data = train_data)
rf_fit  <- fit(rf_wf,  data = train_data)

#| label: fig-ao4-gains
#| fig-cap: "Gains chart — LDA predicted probability deciles vs. actual high-value rate"

lda_preds <- augment(lda_fit, new_data = test_data)
rf_preds  <- augment(rf_fit,  new_data = test_data)

eval_metrics <- metric_set(roc_auc, sens, yardstick::spec)
bind_rows(
  lda_preds |> eval_metrics(truth = high_value, .pred_1, estimate = .pred_class) |> mutate(model = "LDA"),
  rf_preds  |> eval_metrics(truth = high_value, .pred_1, estimate = .pred_class) |> mutate(model = "Random Forest")
) |> select(model, .metric, .estimate) |> pivot_wider(names_from = .metric, values_from = .estimate)

lda_preds |>
  mutate(decile = ntile(desc(.pred_1), 10)) |>
  group_by(decile) |>
  summarise(actual_rate = mean(high_value == "1"), .groups = "drop") |>
  ggplot(aes(x = factor(decile), y = actual_rate)) +
  geom_col(fill = "#154734") +
  labs(title = "Gains Chart — LDA", x = "Predicted Probability Decile (1 = highest)", y = "Actual High-Value Rate") +
  theme_minimal(base_size = 11)
```

# Interpretation

-   Class-conditional means: high-value customers' first orders average
    higher price, lower discount than non-high-value
-   PDP: predicted probability rises with price, falls with discount
    depth — both roughly linear
-   Random Forest variable importance confirms `product_price` \>
    `discount_percent`, consistent with LDA's separating structure

# Summary Findings

-   LDA = best model, ROC AUC = 0.679 vs. Random Forest 0.590 — a real,
    meaningful gap (not within-noise like AO3's model comparison)
-   Acquisition channel and product category showed **no** relationship
    to customer value in this dataset (p = 0.15, p = 0.41) — diverges
    from original hypothesis, which assumed channel mattered
-   First-order price and discount depth are the actionable,
    pre-purchase levers driving predicted customer value
-   Deployed as a probability-scoring model (`high_value_model.joblib` +
    `predict.py`) — customer's first-order price/discount in, high-value
    probability out, business sets its own targeting cutoff

# Analytics Objective 5 — Sales Trends & Seasonality

**Objective:** Determine how sales performance changes over time and whether seasonal patterns differ across product categories.

**Hypotheses:**

- Revenue and order volume will vary across months and quarters.
- Seasonal sales patterns will differ across product categories.
- Identifying peak sales periods can improve inventory planning, promotional timing, and marketing strategy.

```{r ao5-setup}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(lubridate)
library(ranger)
library(xgboost)

set.seed(2026)

ao5_raw <- readr::read_csv(
  "shopify_sales_dataset_ml_eda 2.csv",
  show_col_types = FALSE
) |>
  mutate(
  order_date = lubridate::ymd(order_date)
) |>
  filter(
    !is.na(order_date),
    !is.na(revenue),
    !is.na(quantity),
    !is.na(product_category)
  )

ao5_monthly_category <- ao5_raw |>
  mutate(
    order_month = floor_date(order_date, "month"),
    year = year(order_date),
    month_num = month(order_date),
    month_name = factor(
      month_num,
      levels = 1:12,
      labels = month.abb
    ),
    quarter = factor(
      paste0(
        "Q",
        lubridate::quarter(order_date)
      ),
      levels = paste0("Q", 1:4)
    ),
    product_category = factor(product_category)
  ) |>
  group_by(
    order_month,
    year,
    month_num,
    month_name,
    quarter,
    product_category
  ) |>
  summarise(
    monthly_revenue = sum(revenue, na.rm = TRUE),
    quantity_sold = sum(quantity, na.rm = TRUE),
    order_volume = n(),
    .groups = "drop"
  ) |>
  arrange(order_month, product_category) |>
  mutate(
    time_index = match(
      order_month,
      sort(unique(order_month))
    )
  )

ao5_monthly_overall <- ao5_raw |>
  mutate(
    order_month = floor_date(order_date, "month")
  ) |>
  group_by(order_month) |>
  summarise(
    monthly_revenue = sum(revenue, na.rm = TRUE),
    quantity_sold = sum(quantity, na.rm = TRUE),
    order_volume = n(),
    .groups = "drop"
  ) |>
  mutate(
    month_name = factor(
      month(order_month),
      levels = 1:12,
      labels = month.abb
    ),
    quarter = factor(
      paste0(
        "Q",
        lubridate::quarter(order_month)
      ),
      levels = paste0("Q", 1:4)
    )
  )
```

# Machine Learning Methods

- **Statistical Baseline:** One-way ANOVA and two-way ANOVA
- **Regression:** Linear Regression
- **Ensemble Models:** Random Forest and XGBoost
- **Primary Outcome:** Monthly revenue by product category
- **Secondary Performance Measures:** Quantity sold and order volume
- **Validation:** Time-based holdout rather than a random split

The models use **month, product category, and time trend** to predict monthly revenue.

Quarter is evaluated separately through ANOVA because month already captures the underlying calendar seasonality in the predictive models.

# Data Wrangling & Visualization

- Convert `order_date` into a valid date variable
- Extract month, quarter, and year
- Aggregate transactions into monthly sales performance
- Calculate:
  - Monthly revenue
  - Quantity sold
  - Order volume
- Compare trends from January 2023 through June 2025

```{r ao5-monthly-trends}
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

ao5_monthly_overall |>
  select(
    order_month,
    monthly_revenue,
    quantity_sold
  ) |>
  pivot_longer(
    cols = c(
      monthly_revenue,
      quantity_sold
    ),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(
      metric,
      monthly_revenue = "Monthly Revenue ($)",
      quantity_sold = "Quantity Sold"
    )
  ) |>
  ggplot(
    aes(
      x = order_month,
      y = value
    )
  ) +
  geom_line(
    linewidth = 1,
    color = "#154734"
  ) +
  geom_point(
    size = 2,
    color = "#154734"
  ) +
  facet_wrap(
    ~metric,
    scales = "free_y",
    ncol = 1
  ) +
  scale_x_date(
    date_breaks = "3 months",
    date_labels = "%b\n%Y"
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Shopify Sales Performance Over Time",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12)
```

# Seasonality by Product Category

To evaluate whether seasonal patterns differ by product category, average monthly revenue is compared across month × category combinations.

The heatmap uses the two complete calendar years, **2023–2024**, so that the partial 2025 year does not distort seasonal comparisons.

```{r ao5-seasonality-heatmap}
#| echo: false
#| fig-width: 10
#| fig-height: 5

ao5_seasonal_heatmap <- ao5_monthly_category |>
  filter(
    order_month < as.Date("2025-01-01")
  ) |>
  group_by(
    month_name,
    product_category
  ) |>
  summarise(
    avg_monthly_revenue =
      mean(monthly_revenue),
    .groups = "drop"
  )

ao5_seasonal_heatmap |>
  ggplot(
    aes(
      x = month_name,
      y = product_category,
      fill = avg_monthly_revenue
    )
  ) +
  geom_tile(
    color = "white",
    linewidth = 0.5
  ) +
  scale_fill_gradient(
    low = "#dcefe5",
    high = "#154734",
    labels = scales::dollar
  ) +
  labs(
    title = "Average Monthly Revenue by Product Category",
    subtitle = "Complete calendar years: 2023–2024",
    x = "Month",
    y = "Product Category",
    fill = "Avg. Revenue"
  ) +
  theme_minimal(base_size = 12)
```

# Model Fitting

Because this objective involves time, a **time-based train/test split** is more appropriate than randomly mixing earlier and later observations.

- **Training:** January 2023 – December 2024
- **Testing:** January 2025 – June 2025
- Predictors:
  - Month
  - Product category
  - Time trend
- Models:
  - Linear Regression
  - Random Forest
  - XGBoost

```{r ao5-model-fitting}
#| include: false
#| message: false
#| warning: false
#| cache: true

ao5_model_data <- ao5_monthly_category |>
  select(
    order_month,
    monthly_revenue,
    month_name,
    product_category,
    time_index
  )

ao5_train <- ao5_model_data |>
  filter(
    order_month < as.Date("2025-01-01")
  )

ao5_test <- ao5_model_data |>
  filter(
    order_month >= as.Date("2025-01-01")
  )

ao5_recipe <- recipe(
  monthly_revenue ~
    month_name +
    product_category +
    time_index,
  data = ao5_train
) |>
  step_unknown(
    all_nominal_predictors()
  ) |>
  step_dummy(
    all_nominal_predictors(),
    one_hot = TRUE
  ) |>
  step_zv(
    all_predictors()
  )

ao5_lm_spec <- linear_reg() |>
  set_engine("lm")

ao5_rf_spec <- rand_forest(
  trees = 500,
  mtry = 5,
  min_n = 5
) |>
  set_engine(
    "ranger",
    importance = "impurity"
  ) |>
  set_mode("regression")

ao5_xgb_spec <- boost_tree(
  trees = 300,
  tree_depth = 4,
  learn_rate = 0.05,
  min_n = 5
) |>
  set_engine("xgboost") |>
  set_mode("regression")

ao5_lm_wf <- workflow() |>
  add_recipe(ao5_recipe) |>
  add_model(ao5_lm_spec)

ao5_rf_wf <- workflow() |>
  add_recipe(ao5_recipe) |>
  add_model(ao5_rf_spec)

ao5_xgb_wf <- workflow() |>
  add_recipe(ao5_recipe) |>
  add_model(ao5_xgb_spec)

ao5_lm_fit <- fit(
  ao5_lm_wf,
  data = ao5_train
)

ao5_rf_fit <- fit(
  ao5_rf_wf,
  data = ao5_train
)

ao5_xgb_fit <- fit(
  ao5_xgb_wf,
  data = ao5_train
)

ao5_anova_data <- ao5_monthly_category |>
  filter(
    order_month <
      as.Date("2025-01-01")
  )

ao5_quarter_anova <- aov(
  monthly_revenue ~ quarter,
  data = ao5_anova_data
)

ao5_interaction_anova <- aov(
  monthly_revenue ~
    quarter * product_category,
  data = ao5_anova_data
)

ao5_quarter_p <-
  summary(
    ao5_quarter_anova
  )[[1]][["Pr(>F)"]][1]

ao5_interaction_p <-
  summary(
    ao5_interaction_anova
  )[[1]][["Pr(>F)"]][3]
```

# Prediction & Model Evaluation

Models are evaluated on **January–June 2025**, which was not used during training.

- **RMSE:** Lower is better
- **MAE:** Lower is better
- **R²:** Higher is better

```{r ao5-evaluation}
#| echo: false
#| message: false
#| warning: false
#| cache: true

ao5_metrics_function <- metric_set(
  rmse,
  mae,
  rsq
)

ao5_lm_predictions <- augment(
  ao5_lm_fit,
  new_data = ao5_test
)

ao5_rf_predictions <- augment(
  ao5_rf_fit,
  new_data = ao5_test
)

ao5_xgb_predictions <- augment(
  ao5_xgb_fit,
  new_data = ao5_test
)

ao5_metrics <- bind_rows(

  ao5_lm_predictions |>
    ao5_metrics_function(
      truth = monthly_revenue,
      estimate = .pred
    ) |>
    mutate(
      model = "Linear Regression"
    ),

  ao5_rf_predictions |>
    ao5_metrics_function(
      truth = monthly_revenue,
      estimate = .pred
    ) |>
    mutate(
      model = "Random Forest"
    ),

  ao5_xgb_predictions |>
    ao5_metrics_function(
      truth = monthly_revenue,
      estimate = .pred
    ) |>
    mutate(
      model = "XGBoost"
    )

) |>
  select(
    model,
    .metric,
    .estimate
  )

ao5_metrics |>
  mutate(
    .estimate = round(
      .estimate,
      3
    )
  ) |>
  pivot_wider(
    names_from = .metric,
    values_from = .estimate
  ) |>
  select(
    model,
    rmse,
    mae,
    rsq
  ) |>
  knitr::kable(
    col.names = c(
      "Model",
      "RMSE",
      "MAE",
      "R²"
    ),
    caption =
      "AO5 Performance on 2025 Holdout Data"
  )
```

# Model Comparison

```{r ao5-model-comparison}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

ao5_metrics |>
  filter(
    .metric == "rmse"
  ) |>
  ggplot(
    aes(
      x = reorder(
        model,
        .estimate
      ),
      y = .estimate
    )
  ) +
  geom_col(
    fill = "#154734",
    alpha = 0.85
  ) +
  geom_text(
    aes(
      label = scales::comma(
        round(.estimate)
      )
    ),
    hjust = -0.15,
    size = 4
  ) +
  coord_flip() +
  labs(
    title =
      "Monthly Revenue Prediction — Model Comparison",
    subtitle =
      "Lower test RMSE indicates better performance",
    x = NULL,
    y = "Test RMSE ($)"
  ) +
  theme_minimal(base_size = 12)
```

# Interpretation

Random Forest variable importance provides an additional diagnostic view of which calendar and category features contribute the most information to monthly revenue predictions.

These importance scores indicate **predictive signal**, not causation.

```{r ao5-importance}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| cache: true

ao5_rf_engine <- ao5_rf_fit |>
  extract_fit_engine()

ao5_importance <- tibble(
  Variable = names(
    ao5_rf_engine$variable.importance
  ),
  Importance = as.numeric(
    ao5_rf_engine$variable.importance
  )
) |>
  arrange(
    desc(Importance)
  )

ao5_importance |>
  slice_head(
    n = 10
  ) |>
  ggplot(
    aes(
      x = reorder(
        Variable,
        Importance
      ),
      y = Importance
    )
  ) +
  geom_col(
    fill = "#154734",
    alpha = 0.85
  ) +
  coord_flip() +
  labs(
    title =
      "Feature Importance — Sales Trends & Seasonality",
    x = "Predictor",
    y = "Importance"
  ) +
  theme_minimal(base_size = 12)
```

```{r ao5-summary-values}
#| include: false

ao5_best_model <- ao5_metrics |>
  filter(
    .metric == "rmse"
  ) |>
  slice_min(
    .estimate,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(model)

ao5_best_rmse <- ao5_metrics |>
  filter(
    model == ao5_best_model,
    .metric == "rmse"
  ) |>
  pull(.estimate)

ao5_best_mae <- ao5_metrics |>
  filter(
    model == ao5_best_model,
    .metric == "mae"
  ) |>
  pull(.estimate)

ao5_best_rsq <- ao5_metrics |>
  filter(
    model == ao5_best_model,
    .metric == "rsq"
  ) |>
  pull(.estimate)

ao5_complete_years <- ao5_monthly_overall |>
  filter(
    order_month <
      as.Date("2025-01-01")
  )

ao5_peak_month_table <-
  ao5_complete_years |>
  group_by(month_name) |>
  summarise(
    avg_monthly_revenue =
      mean(monthly_revenue),
    .groups = "drop"
  )

ao5_peak_month <-
  ao5_peak_month_table |>
  slice_max(
    avg_monthly_revenue,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(month_name)

ao5_peak_month_revenue <-
  ao5_peak_month_table |>
  slice_max(
    avg_monthly_revenue,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(avg_monthly_revenue)

ao5_peak_quarter_table <-
  ao5_complete_years |>
  group_by(quarter) |>
  summarise(
    avg_monthly_revenue =
      mean(monthly_revenue),
    .groups = "drop"
  )

ao5_peak_quarter <-
  ao5_peak_quarter_table |>
  slice_max(
    avg_monthly_revenue,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(quarter)

ao5_category_summary <-
  ao5_monthly_category |>
  filter(
    order_month <
      as.Date("2025-01-01")
  ) |>
  group_by(
    product_category
  ) |>
  summarise(
    avg_monthly_revenue =
      mean(monthly_revenue),
    .groups = "drop"
  )

ao5_top_category <-
  ao5_category_summary |>
  slice_max(
    avg_monthly_revenue,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(product_category)

ao5_top_feature <-
  ao5_importance |>
  slice_max(
    Importance,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(Variable)
```

# Summary Findings

- **Best Model:** `r ao5_best_model`
- **Test RMSE:** $`r scales::comma(round(ao5_best_rmse))`
- **Test MAE:** $`r scales::comma(round(ao5_best_mae))`
- **Test R²:** `r round(ao5_best_rsq, 3)`
- **Peak Month:** `r ao5_peak_month` — average monthly revenue of approximately $`r scales::comma(round(ao5_peak_month_revenue))`
- **Peak Quarter:** `r ao5_peak_quarter`
- **Highest Average-Revenue Category:** `r ao5_top_category`
- **Top Random Forest Feature:** `r ao5_top_feature`

**Quarter Effect:** `r if (ao5_quarter_p < 0.05) paste0("Quarterly differences were statistically significant (p = ", round(ao5_quarter_p, 3), ").") else paste0("Quarterly differences were not statistically significant (p = ", round(ao5_quarter_p, 3), ").")`

**Quarter × Product Category Interaction:** `r if (ao5_interaction_p < 0.05) paste0("Seasonal patterns differed significantly across product categories (p = ", round(ao5_interaction_p, 3), ").") else paste0("The quarter × product category interaction was not statistically significant (p = ", round(ao5_interaction_p, 3), ").")`

`r if (ao5_best_rsq < 0.10) "Overall predictive strength is weak, suggesting that month, product category, and historical time trend alone do not explain much of the variation in future monthly revenue." else if (ao5_best_rsq < 0.30) "The models show modest predictive strength, indicating that seasonality and product category provide some forecasting information but additional variables would improve predictions." else "The results indicate that calendar seasonality and product category provide meaningful information for forecasting monthly revenue."`

Businesses should use historical seasonal patterns to support **inventory, promotional, and marketing planning**, while recognizing that calendar timing alone may not be sufficient for accurate sales forecasting.

# Conclusion

# Key Findings

-   **Marketing Channel Performance (AO1):** Marketing channels vary in
    their contribution to sales and profitability, showing that
    transaction volume alone does not provide a complete picture of
    channel performance

-   **Customer Satisfaction (AO2):** Customer ratings can be analyzed
    alongside product, pricing, discount, and purchasing characteristics
    to identify factors associated with satisfaction

-   **Product Returns (AO3):** Return behavior can be predicted using
    customer and order characteristics, although class imbalance makes
    identifying returned orders more challenging

-   **Customer Value & Retention (AO4):** Repeat purchasing behavior
    helps distinguish between customers generating short-term sales and
    those contributing to longer-term customer value

-   **Seasonality (AO5):** Sales performance changes over time,
    demonstrating the importance of seasonal patterns when forecasting
    demand and planning marketing activity

-   **Overall Finding:** Profitability cannot be explained by sales
    volume alone — acquisition source, satisfaction, returns, repeat
    purchasing, discounts, and seasonality together provide a more
    complete picture of business performance **:::**

# Recommendations for Businesses

- **Optimize marketing spending by channel and product category**
  - Compare traffic sources based on the average order value they generate across product categories.
  - Prioritize channel × category combinations that consistently produce stronger order values.

- **Use customer satisfaction as an early performance signal**
  - Monitor customer ratings alongside product category, traffic source, discounts, and purchasing behavior to identify areas where customer satisfaction may be improved.

- **Improve return-risk data collection**
  - Shipping cost and discount percentage showed the strongest relative signals within the current return model.
  - Because predictive performance was weak, collect additional customer, fulfillment, and behavioral data before using return-risk predictions operationally.

- **Strengthen customer retention efforts**
  - Compare repeat and one-time customers based on cumulative profit, return behavior, and acquisition source.
  - Prioritize retention strategies when repeat customers demonstrate stronger long-term profitability than continued reliance on new-customer acquisition.

- **Use seasonal patterns to improve planning**
  - Use observed monthly and quarterly sales patterns to support inventory, promotional, and marketing decisions.
  - Focus planning around the strongest periods and product-category patterns identified in the analysis.

# Limitations and Directions for Future Research

::::::: columns
:::: {.column width="50%"}
### Limitations

::: incremental
-   Kaggle notes the dataset is "realistic" rather than verified,
    meaning it is likely **synthetic** — findings may not generalize to
    real Shopify transaction data
-   Findings are scoped to this dataset's countries, categories, and
    date range (Jan 2023–Jun 2025) and may not generalize beyond it
-   Each Analytics Objective models its outcome independently — none of
    the five models account for interactions *across* objectives (e.g.,
    a channel's effect on both satisfaction and return risk together)
-   Several models (e.g., AO3's return prediction) face class imbalance,
    which caps achievable sensitivity even after optimizing evaluation
    metrics like ROC AUC
-   The deployed model reflects a single static training snapshot and
    does not retrain or adapt as new orders come in
:::
::::

:::: {.column width="50%"}
### Future Research

::: incremental
-   Validate findings against real, non-synthetic transaction data
    before informing operational decisions
-   Build an integrated model connecting acquisition, satisfaction,
    returns, and retention rather than five separate models, to capture
    cross-objective interactions
-   Extend the population beyond the current
:::
::::
:::::::

# Appendix

-   [GitHub Repo](https://github.com/lewiswaddell/GP-W10.git)

-   [Presentation Slides](https://lewiswaddell.github.io/GP-W10/)

-   [Deploy Model Shiny App AO3](https://ibm-6540-project.shinyapps.io/model-deployment/)
