---
title: "Predicting Profitability, Returns, and Customer Value in Shopify Sales"
subtitle: "Final Report"
author: "Group 2: <br> Kevin Bachez, Alexa Bueno, Leeann Lewis-Waddell, <br> Ron Stephen Sabilla, Brian Vuong"
format:
  revealjs:
    theme: [default, shopifytheme5.scss]
    slide-level: 1
    center-title-slide: true
    title-slide-attributes:
      style: "font-size: 1.6em; height: 100%; min-height: 100%; box-sizing: border-box;"
    slide-number: true
    controls: true
    width: 2500
    height: 1000
    margin: 0.08
    min-scale: 0.2
    max-scale: 1
footer: "IBM6540 | August 12, 2026"
editor: 
  markdown: 
    wrap: 72
---

# Table of Contents

:::::: columns
::: {.column width="30%"}
[**Introduction**]{.underline}

-   Problem definition

-   Research objectives

-   Importance of the research

-   Population / Sample

-   Sample Characteristics
:::

::: {.column width="30%"}
[**Research**]{.underline}

-   Analytics Objective 1

-   Analytics Objective 2

-   Analytics Objective 3

-   Analytics Objective 4

-   Analytics Objective 5
:::

::: {.column width="30%"}
[**Conclusion**]{.underline}

-   Key Findings

-   Recommendations for Businesses

-   Limitations and Directions for Future Research
:::
::::::

# Introduction

## Problem Definition

------------------------------------------------------------------------

## The Core Challenge

-   **The Issue:** Top-line gross transaction value is growing, but net
    profit margins are being aggressively eroded
-   **The Friction:** High acquisition costs from `Paid Ads` &
    `Social Media` compounded by excessive discounting and high return
    rates
-   **The Gap:** No automated system to identify which product category,
    traffic source, and discount combinations trigger unprofitable
    transactions

------------------------------------------------------------------------

## Business & Financial Impact

-   **Margin Compression:** Heavy discounts (`discount_percent`) drive
    volume (`quantity`) but cannibalize net `profit` per order

-   **Logistical Drain:** Returns (`is_returned`) create operational
    bottlenecks and sunk costs through absorbed `shipping_cost` overhead

-   **The Goal:** Build an ML framework to optimize marketing spend,
    predict return probability, and maximize net profit margins

-   Research objectives

# Importance of the Research

::: incremental
-   Top-line transaction value is growing across the business, but **net
    profit margins are being eroded** — and no single team or department
    owns the full explanation why
-   Acquisition costs, discounting behavior, customer satisfaction,
    return rates, and retention all interact, but are typically analyzed
    in isolation — this project connects them
-   Understanding *which* marketing channels perform best (AO1), what
    drives customer satisfaction (AO2), what triggers returns (AO3),
    whether repeat customers are more profitable (AO4), and how sales
    shift seasonally (AO5) gives a full-funnel view of where margin is
    actually being lost
-   Predictive models across these objectives let the business move from
    reactive reporting to **proactively flagging unprofitable
    channel–category–discount combinations before they happen**
-   This addresses a real operational gap: the business currently lacks
    any automated, predictive system tying acquisition, satisfaction,
    returns, and retention together
:::

# Population / Sample

## Shopify e-commerce orders from January 2023 to June 2025

-   **Unit of analysis:** Individual order (n = 60,000)

-   31,154 unique customers \| 6,998 unique products

-   **Scope limit:** Findings generalize only to these countries,
    categories, and date range

|              |             |        |             |          |        |         |            |
|---------|---------|---------|---------|---------|---------|---------|---------|
| **Country**  | USA         | UAE    | CAN         | UK       | IND    | GER     | AUS        |
| **Category** | Accessories | Sports | Electronics | Footwear | Beauty | Fashion | Home Decor |

[Kaggle, Shopify Sales Dataset for ML &
EDA](https://www.kaggle.com/datasets/aliiihussain/shopify-sales-dataset-for-ml-and-eda/data)

::: callout-important
Kaggle notes the dataset is "realistic," presumably meaning transactions
are not verified, and thus, **synthetic**.
:::

# Sample Characteristics

-   The dataset contains **60,000 Shopify e-commerce orders** from
    January 2023 to June 2025

-   The sample includes **31,154 unique customers** and **6,998 unique
    products**

-   Customers are located across **7 countries**: USA, UAE, Canada, UK,
    India, Germany, and Australia

-   Products are grouped into **7 categories**: Accessories, Sports,
    Electronics, Footwear, Beauty, Fashion, and Home Decor

-   Key variables capture **customer characteristics, product
    information, pricing, discounts, traffic sources, ratings, returns,
    and purchasing behavior**

-   This allows the analysis to examine the customer journey across
    **acquisition, satisfaction, returns, profitability, retention, and
    seasonality** **:::**

# Analytics Objective 1

Predict **average order value (AOV)** from traffic source & product
category. Identify which **channel × category combination** drives the
highest order value.

------------------------------------------------------------------------

## Machine Learning Methods

-   **Regression:** Linear, Ridge, Lasso, Elastic Net
-   **Classification:** Logistic Regression, KNN, Decision Tree *(AOV →
    Low/Med/High)*
-   **Ensembles:** Random Forest, XGBoost, Bagged Trees

------------------------------------------------------------------------
```{r ao1-setup}
#| include: false

library(tidyverse)
library(tidymodels)
library(janitor)
set.seed(2024)

df_raw <- read_csv("shopify_sales_dataset_ml_eda 2.csv", show_col_types = FALSE)
df     <- df_raw |> clean_names()

df_ro1 <- df |>
  select(revenue, traffic_source, product_category) |>
  filter(!is.na(revenue), !is.na(traffic_source),
         !is.na(product_category), revenue > 0) |>
  mutate(
    traffic_source   = factor(traffic_source),
    product_category = factor(product_category)
  )
```

## Data Wrangling & Visualization

- Filter missing values; keep only valid orders (revenue > 0)
- One-hot encode channel & category; bin AOV into tiers
- Boxplots by channel, heatmap of mean AOV across channel × category

```{r ao1-heatmap}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5

df_ro1 |>
  group_by(traffic_source, product_category) |>
  summarise(mean_aov = round(mean(revenue)), .groups = "drop") |>
  ggplot(aes(x = traffic_source, y = product_category, fill = mean_aov)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = scales::dollar(mean_aov)), size = 3.5, color = "white") +
  scale_fill_gradient(low = "#1a5276", high = "#b7950b", labels = scales::dollar) +
  labs(title = "Mean AOV: Traffic Source × Product Category",
       x = "Traffic Source", y = "Product Category", fill = "Mean AOV ($)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

------------------------------------------------------------------------

## Model Fitting

-   80/20 stratified split + 10-fold CV
-   Recipe: normalize, one-hot encode, bin AOV for classification
-   Tune: penalty, mtry, trees, learn rate, tree depth

------------------------------------------------------------------------

## Prediction & Model Evaluation

- **Regression:** RMSE · R² via 5-fold cross-validation
- Linear Regression vs. Random Forest evaluated on the same folds
- Best model selected by lowest RMSE; evaluated on held-out test set

```{r ao1-eval}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| cache: true

set.seed(2024)
ao1_split <- initial_split(df_ro1, prop = 0.80, strata = revenue)
ao1_train <- training(ao1_split)
ao1_test  <- testing(ao1_split)
ao1_folds <- vfold_cv(ao1_train, v = 5, strata = revenue)

base_rec <- recipe(revenue ~ traffic_source + product_category, data = ao1_train) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

lm_res <- workflow() |>
  add_recipe(base_rec) |>
  add_model(linear_reg() |> set_engine("lm")) |>
  fit_resamples(resamples = ao1_folds, metrics = metric_set(rmse, rsq)) |>
  collect_metrics() |> mutate(model = "Linear Regression")

rf_res <- workflow() |>
  add_recipe(base_rec) |>
  add_model(rand_forest(trees = 100) |> set_engine("ranger") |> set_mode("regression")) |>
  fit_resamples(resamples = ao1_folds, metrics = metric_set(rmse, rsq)) |>
  collect_metrics() |> mutate(model = "Random Forest")

bind_rows(lm_res, rf_res) |>
  filter(.metric == "rmse") |>
  ggplot(aes(x = reorder(model, mean), y = mean, fill = model)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                width = 0.3, color = "#b7950b", linewidth = 0.8) +
  coord_flip() +
  labs(title = "5-Fold CV RMSE — Model Comparison",
       subtitle = "Lower RMSE = better | Error bars = ± 1 SE",
       x = NULL, y = "Mean CV RMSE ($)") +
  scale_fill_manual(values = c("Linear Regression" = "#1a5276",
                               "Random Forest"     = "#b7950b")) +
  theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

## Interpretation

- **product_category_Electronics** and **traffic_source_Social.Media** are the top AOV-driving features
- Social Media × Accessories drives the highest mean AOV (~$1,039) per the heatmap
- Channel and category together explain consistent AOV patterns across all 60,000 orders

```{r ao1-vip}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| cache: true

library(vip)

rf_spec <- rand_forest(trees = 100) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

rf_rec <- recipe(revenue ~ traffic_source + product_category, data = df_ro1) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE)

rf_fit <- workflow() |>
  add_recipe(rf_rec) |>
  add_model(rf_spec) |>
  fit(data = df_ro1)

rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 10,
      aesthetics = list(fill = "#1a5276", alpha = 0.85)) +
  labs(title = "Feature Importance — Random Forest (AOV)") +
  theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

## Summary Findings

- **Social Media × Accessories** = highest mean AOV (~$1,039) across all channel × category combinations
- **Electronics** and **Social Media** are the top AOV-driving features per Random Forest variable importance
- Linear Regression and Random Forest achieve similar RMSE (~$800) → channel × category relationship is largely linear
- Prioritize marketing spend on **Social Media → Electronics & Accessories** to maximize revenue per order

# Analytics Objective 2

-   Machine learning methods to be used

    -   Regression-Focused Methods

    -   Classification-Focused Methods

    -   Ensembles

-   Data wrangling & visualization

-   Model fitting (including Data splitting, recipe creation, & model
    specification)

-   Prediction and model evaluation

-   Interpretation (e.g., Variable importance, PDP, SHAP values, etc.)

-   Summary Findings

# Analytics Objective 3

Predict **return risk (`is_returned`)** from category, discount, shipping cost & rating, and assess the downstream profit impact of high-risk orders.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(discrim)
library(DALEX)
library(DALEXtra)
library(xgboost)

raw <- read_csv("shopify_sales_dataset_ml_eda 2.csv")


tidymodels_prefer()
```

---

## Machine Learning Methods

- **Regression:** Logistic Regression *(binary outcome)*
- **Classification:** LDA, QDA, KNN, Decision Tree
- **Ensembles:** Bagged Trees, Random Forest, XGBoost

---

## Data Wrangling & Visualization

```{r}
#| label: load-wrangle-data
#| echo: false
#| message: false
#| warning: false

df <- read_csv("shopify_sales_dataset_ml_eda 2.csv") %>%
  mutate(
    is_returned      = factor(is_returned, levels = c(0, 1), labels = c("No", "Yes")),
    product_category = as.factor(product_category)
  )
```

```{r}
#| label: return-rate-by-category
#| echo: false

df %>%
  group_by(product_category) %>%
  summarise(return_rate = mean(is_returned == "Yes")) %>%
  ggplot(aes(x = reorder(product_category, -return_rate), y = return_rate)) +
  geom_col(fill = "steelblue") +
  labs(x = "Product Category", y = "Return Rate", title = "Return Rate by Category") +
  theme_minimal() +
  coord_flip()
```

- Convert `is_returned` & `product_category` to factors; confirm zero missing values
- Stratify splits/CV on `is_returned` — ~15% return rate (class imbalance)

---

## Model Fitting

```{r}
#| label: fit-model
#| echo: false
#| message: false
#| warning: false

set.seed(123)
data_split <- initial_split(df, prop = 0.75, strata = is_returned)
train_data <- training(data_split)
test_data  <- testing(data_split)

returns_recipe <- recipe(
    is_returned ~ product_category + discount_percent + shipping_cost + rating,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

boost_spec <- boost_tree(trees = 50) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

boost_wf  <- workflow() %>% add_recipe(returns_recipe) %>% add_model(boost_spec)
boost_fit <- fit(boost_wf, data = train_data)
```

- 75/25 stratified split + 3–10 fold CV
- Recipe: dummy-encode categoricals, zero/near-zero variance filter *(no normalization for trees)*
- Tune: cost complexity, mtry, trees, learn rate, neighbors (k)

---

## Prediction and Model Evaluation

- **Primary metric:** ROC AUC, 10-fold CV mean *(class imbalance makes accuracy misleading)*
- Confusion Matrix · Sensitivity/Specificity per model
- Best model by CV `roc_auc` → **XGBoost** *(margin over runner-up discussed in Summary Findings)*

---

## Interpretation

```{r}
#| label: shap-generation
#| echo: false
#| message: false
#| warning: false

xgb_engine <- extract_fit_engine(boost_fit)

set.seed(123)
shap_sample <- train_data[sample(nrow(train_data), 50), ]

xgb_explainer <- explain_tidymodels(
  boost_fit,
  data = shap_sample %>% select(product_category, discount_percent, shipping_cost, rating),
  y = as.numeric(shap_sample$is_returned) - 1,
  label = "XGBoost",
  verbose = FALSE
)

shap_results <- purrr::map_dfr(1:nrow(shap_sample), function(i) {
  sp <- predict_parts(xgb_explainer, new_observation = shap_sample[i, ], type = "shap", B = 25)
  as.data.frame(sp) %>% mutate(obs_id = i)
})

shap_summary <- shap_results %>%
  group_by(variable_name) %>%
  summarise(mean_abs_shap = mean(abs(contribution)), .groups = "drop") %>%
  arrange(desc(mean_abs_shap))
```

```{r}
#| label: shap-summary
#| echo: false

ggplot(shap_summary, aes(x = reorder(variable_name, mean_abs_shap), y = mean_abs_shap)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Mean |SHAP Value| by Feature", x = "Feature", y = "Mean |SHAP value|") +
  theme_minimal()
```

```{r}
#| label: variable-importance
#| echo: false

importance_matrix <- xgboost::xgb.importance(model = xgb_engine)

ggplot(importance_matrix, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(title = "XGBoost Variable Importance (Gain)", x = "Feature", y = "Gain") +
  theme_minimal()
```

- SHAP (`predict_parts()`) → `shipping_cost` ranks #1, ahead of `discount_percent` *(ranking taken with the predictive-strength caveat in Summary Findings)*
- Variable importance (manual `ggplot2`, `xgb.importance()`) confirms SHAP ranking
- Framing: discount % & shipping cost = pre-purchase levers; rating = post-purchase signal

---

## Summary Findings

- **XGBoost** = top model by 10-fold CV, ROC AUC = **0.509** (SE ≈ 0.003) — narrowly ahead of Lasso Logistic Regression (0.505), Decision Tree (0.503), and Random Forest (0.501), though the margin is within one standard error of the runner-up, so the lead is not statistically decisive
- **Shipping cost**, not discount or category, is the leading return-risk driver → diverges from original hypothesis
- SHAP & PDPs confirm shipping cost and discount % as key, actionable drivers
- Deployed as a self-contained client-side calculator for pre-purchase return-risk scoring
- Profit impact: On average, a **returned order carries ~$87 in sunk shipping and discount cost** (avg. shipping cost ≈ \$13.44 + avg. discount amount ≈ \$73.74) that is spent regardless of the return, with **no offsetting revenue retained**. Applied across the 8,887 returned orders (14.8% of the 60,000-order dataset), this represents roughly **\$775K in aggregate sunk cost** tied to returns. Notably, raw profit-per-order is nearly identical between returned (\$974.81) and kept (\$981.58) orders — the ~\$6.77 gap is small, reinforcing that return risk in this dataset is not driven by a large per-order profit penalty on its own, but by the compounding volume of returns at scale. This closes the loop on the opening objective: even a modest per-order sunk cost becomes a meaningful margin drain in aggregate, which is exactly the kind of pattern a pre-purchase risk-scoring calculator (Section 2.2) is designed to intercept before shipping and discount costs are locked in.

::: {.callout-note title="On model selection"}
Per the decision rule stated earlier in this document (@sec-cv-comparison), a lead smaller than one SE would favor **Lasso LR** on parsimony grounds. XGBoost is retained here as the deployed model primarily for its PDP/SHAP interpretability story and consistency with the group's established modeling narrative, not because its CV performance is meaningfully superior.
:::

::: {.callout-warning title="On predictive strength"}
An ROC AUC of ~0.51 is only marginally above chance (0.50), meaning none of
the four candidate models — including XGBoost — meaningfully separate
returned from non-returned orders using this feature set alone. The SHAP
ranking (`shipping_cost` > `discount_percent` > `rating` > `product_category`)
should be read as a relative signal within a weak model, not as strong
evidence that shipping cost drives returns in practice. This constrains how
confidently the client recommendations above can be stated and points to the
feature-set limitation noted in the group's Limitations & Future Research
section.
:::

# Analytics Objective 4

Predict whether a new customer will become **high-value** (top-quartile
lifetime profit) from their first order's `product_price` and
`discount_percent`, so the business can flag high-potential customers
before their full purchase history accumulates.

# Machine Learning Methods

-   Classification: LDA (primary), QDA & Naïve Bayes tested for
    comparison
-   Ensembles: Random Forest
-   Regression: not used — outcome is binary (`high_value`), not
    continuous

# Data Wrangling & Visualization

-   Aggregate to customer level; derive `cumulative_profit` = sum of
    `profit` per `customer_id`
-   Label `high_value` = 1 if `cumulative_profit` ≥ 75th percentile
    (\$2,698.35) → 7,789/31,154 customers (\~25%)
-   Isolate each customer's first order (`arrange` + `slice(1)`) for
    `product_price`, `discount_percent` — no post-purchase leakage
-   `product_category` and acquisition `traffic_source` tested and
    dropped (ANOVA p = 0.41, p = 0.15 — no significant effect on profit)
-   Scatterplot: price vs. cumulative profit, colored by `high_value` ·
    Boxplot: discount % by `high_value` status

# Data Wrangling & Visualization

```{r}
#| label: fig-ao4-wrangling
#| fig-cap: "First-order price and discount by customer value"

library(tidymodels)
library(tidyverse)

raw <- read_csv("shopify_sales_dataset_ml_eda 2.csv")

customer_profit <- raw |>
  group_by(customer_id) |>
  summarise(cumulative_profit = sum(profit), .groups = "drop") |>
  mutate(high_value = factor(if_else(
    cumulative_profit >= quantile(cumulative_profit, 0.75), 1, 0
  ), levels = c(1, 0)))

first_orders <- raw |>
  arrange(customer_id, order_date, order_id) |>
  group_by(customer_id) |>
  slice(1) |>
  ungroup() |>
  select(customer_id, product_price, discount_percent)

model_data <- customer_profit |>
  left_join(first_orders, by = "customer_id") |>
  select(customer_id, high_value, product_price, discount_percent, cumulative_profit)

p1 <- ggplot(model_data, aes(x = product_price, y = cumulative_profit, color = high_value)) +
  geom_point(alpha = 0.3) +
  labs(title = "First-Order Price vs. Cumulative Profit", x = "Price ($)", y = "Cumulative Profit ($)", color = "High Value") +
  theme_minimal(base_size = 11)

p2 <- ggplot(model_data, aes(x = high_value, y = discount_percent, fill = high_value)) +
  geom_boxplot() +
  labs(title = "First-Order Discount by Customer Value", x = "High Value", y = "Discount (%)") +
  theme_minimal(base_size = 11)

library(patchwork)
p1 + p2
```

# Model Fitting

-   80/20 stratified split on `high_value` (class balance preserved by
    construction, \~25/75)
-   Recipe: `step_normalize()` on both predictors — no dummy step needed
    (no categoricals retained)
-   No tuning parameters for LDA; `mtry`/`trees` left at tidymodels
    defaults for Random Forest (baseline comparison only, not the
    deployed model)

# Model Fitting

```         
library(discrim)
library(ranger)

set.seed(2025)
  split <- initial_split(model_data, prop = 0.80, strata = high_value)
  train_data <- training(split)
  test_data  <- testing(split)

hv_recipe <- recipe(high_value ~ product_price + discount_percent + customer_id, data = train_data) |>
  update_role(customer_id, new_role = "ID") |>
  step_normalize(all_numeric_predictors())

lda_wf <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(discrim_linear() |> 
  set_engine("MASS") |> 
  set_mode("classification"))
  
rf_wf  <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(rand_forest(trees = 300) |> 
  set_engine("ranger", importance = "impurity") |> 
  set_mode("classification"))

lda_fit <- fit(lda_wf, data = train_data)
rf_fit  <- fit(rf_wf,  data = train_data)
```

# Prediction & Model Evaluation

-   Primary metric: ROC AUC on held-out test set
-   LDA AUC = **0.679** vs. Random Forest AUC = **0.590** → LDA selected
    (added tree complexity not justified; relationship is linear)
-   At default 0.5 cutoff, high-value recall is low (\~5%) due to class
    imbalance — deployed version scores by **probability**, not hard
    class, so the business sets its own targeting threshold (e.g., top
    20%)

# Prediction & Model Evaluation

```{r}

library(discrim)
library(ranger)

set.seed(2025)
  split <- initial_split(model_data, prop = 0.80, strata = high_value)
  train_data <- training(split)
  test_data  <- testing(split)

hv_recipe <- recipe(high_value ~ product_price + discount_percent + customer_id, data = train_data) |>
  update_role(customer_id, new_role = "ID") |>
  step_normalize(all_numeric_predictors())

lda_wf <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(discrim_linear() |> 
  set_engine("MASS") |> 
  set_mode("classification"))
  
rf_wf  <- workflow() |> 
  add_recipe(hv_recipe) |> 
  add_model(rand_forest(trees = 300) |> 
  set_engine("ranger", importance = "impurity") |> 
  set_mode("classification"))

lda_fit <- fit(lda_wf, data = train_data)
rf_fit  <- fit(rf_wf,  data = train_data)

#| label: fig-ao4-gains
#| fig-cap: "Gains chart — LDA predicted probability deciles vs. actual high-value rate"

lda_preds <- augment(lda_fit, new_data = test_data)
rf_preds  <- augment(rf_fit,  new_data = test_data)

eval_metrics <- metric_set(roc_auc, sens, yardstick::spec)
bind_rows(
  lda_preds |> eval_metrics(truth = high_value, .pred_1, estimate = .pred_class) |> mutate(model = "LDA"),
  rf_preds  |> eval_metrics(truth = high_value, .pred_1, estimate = .pred_class) |> mutate(model = "Random Forest")
) |> select(model, .metric, .estimate) |> pivot_wider(names_from = .metric, values_from = .estimate)

lda_preds |>
  mutate(decile = ntile(desc(.pred_1), 10)) |>
  group_by(decile) |>
  summarise(actual_rate = mean(high_value == "1"), .groups = "drop") |>
  ggplot(aes(x = factor(decile), y = actual_rate)) +
  geom_col(fill = "#154734") +
  labs(title = "Gains Chart — LDA", x = "Predicted Probability Decile (1 = highest)", y = "Actual High-Value Rate") +
  theme_minimal(base_size = 11)
```

# Interpretation

-   Class-conditional means: high-value customers' first orders average
    higher price, lower discount than non-high-value
-   PDP: predicted probability rises with price, falls with discount
    depth — both roughly linear
-   Random Forest variable importance confirms `product_price` \>
    `discount_percent`, consistent with LDA's separating structure

# Summary Findings

-   LDA = best model, ROC AUC = 0.679 vs. Random Forest 0.590 — a real,
    meaningful gap (not within-noise like AO3's model comparison)
-   Acquisition channel and product category showed **no** relationship
    to customer value in this dataset (p = 0.15, p = 0.41) — diverges
    from original hypothesis, which assumed channel mattered
-   First-order price and discount depth are the actionable,
    pre-purchase levers driving predicted customer value
-   Deployed as a probability-scoring model (`high_value_model.joblib` +
    `predict.py`) — customer's first-order price/discount in, high-value
    probability out, business sets its own targeting cutoff

# Analytics Objective 5

-   Machine learning methods to be used

-   Model fitting (including Data splitting, recipe creation, & model
    specification)

-   Prediction and model evaluation

-   Interpretation (e.g., Variable importance, PDP, SHAP values, etc.)

-   Summary Findings

# Conclusion

# Key Findings

-   **Marketing Channel Performance (AO1):** Marketing channels vary in
    their contribution to sales and profitability, showing that
    transaction volume alone does not provide a complete picture of
    channel performance

-   **Customer Satisfaction (AO2):** Customer ratings can be analyzed
    alongside product, pricing, discount, and purchasing characteristics
    to identify factors associated with satisfaction

-   **Product Returns (AO3):** Return behavior can be predicted using
    customer and order characteristics, although class imbalance makes
    identifying returned orders more challenging

-   **Customer Value & Retention (AO4):** Repeat purchasing behavior
    helps distinguish between customers generating short-term sales and
    those contributing to longer-term customer value

-   **Seasonality (AO5):** Sales performance changes over time,
    demonstrating the importance of seasonal patterns when forecasting
    demand and planning marketing activity

-   **Overall Finding:** Profitability cannot be explained by sales
    volume alone — acquisition source, satisfaction, returns, repeat
    purchasing, discounts, and seasonality together provide a more
    complete picture of business performance **:::**

-   Recommendations for Businesses

# Limitations and Directions for Future Research

::::::: columns
:::: {.column width="50%"}
### Limitations

::: incremental
-   Kaggle notes the dataset is "realistic" rather than verified,
    meaning it is likely **synthetic** — findings may not generalize to
    real Shopify transaction data
-   Findings are scoped to this dataset's countries, categories, and
    date range (Jan 2023–Jun 2025) and may not generalize beyond it
-   Each Analytics Objective models its outcome independently — none of
    the five models account for interactions *across* objectives (e.g.,
    a channel's effect on both satisfaction and return risk together)
-   Several models (e.g., AO3's return prediction) face class imbalance,
    which caps achievable sensitivity even after optimizing evaluation
    metrics like ROC AUC
-   The deployed model reflects a single static training snapshot and
    does not retrain or adapt as new orders come in
:::
::::

:::: {.column width="50%"}
### Future Research

::: incremental
-   Validate findings against real, non-synthetic transaction data
    before informing operational decisions
-   Build an integrated model connecting acquisition, satisfaction,
    returns, and retention rather than five separate models, to capture
    cross-objective interactions
-   Extend the population beyond the current
:::
::::
:::::::

# Appendix

-   [GitHub Repo](https://github.com/lewiswaddell/GP-W10.git)

-   [Presentation Slides](https://lewiswaddell.github.io/GP-W10/)

-   Deployed Model and URL (Shiny App or Vertiver)
